Experience Report: Kandria - A Game in Common Lisp

### 2023-02-11 els2023-kandria/paper.tex
Revision

\author{Nicolas Hafner}
\affiliation{%
  \institution{Shirakumo.org}
  \city{Zürich}
  \country{Switzerland}
}
\email{shinmera@tymoon.eu}

@@

\begin{CCSXML}

\end{CCSXML}

\keywords{Common Lisp, Games, Video Games, Computer Graphics, Experience Report}

@@

All of the work we've done, including Trial, and even Kandria, is open source and available on our \href{https://github.com/shirakumo}{GitHub}, in the hopes that it will inspire others to create new projects based upon them.

@@

@@

Usual desktop UI toolkits are rather cumbersome to style effectively, as most desktop applications are expected to present themselves in a ``native look and feel'', while games are expected to have rather elaborately customised and animated interfaces. Alloy's presentations protocol allows us to style the UI quite extensively, with relative ease (\autoref{lst:presentation}).

\begin{listing}
\begin{minted}[fontsize=\small]{common-lisp}
;; Define the shapes that compose the element
(presentations:define-realization (ui alloy:switch)
  ((:border simple:rectangle)
   (alloy:margins)
   :pattern colors:white
   :line-width (alloy:un 2))
  ((:background simple:rectangle)
   (alloy:margins))
  ((:switch simple:rectangle)
   (alloy:extent 0 0 (alloy:pw 0.3) (alloy:ph))))

;; Define how shape properties change
(presentations:define-update (ui alloy:switch)
  (:border
   :hidden-p NIL
   :pattern (if alloy:focus
                (colored:color 0.9 0.9 0.9)
                colors:gray))
  (:background
   :pattern (if (alloy:active-p alloy:renderable)
                colors:dark-accent
                (colored:color 0 0 0 0.5)))
  (:switch
   :pattern (if (alloy:active-p alloy:renderable)
                colors:accent
                colors:black)))

;; Define how to animate the properties
(presentations:define-animated-shapes alloy:switch
  (:background (simple:pattern :duration 0.2))
  (:switch (simple:pattern :duration 0.3)
           (simple:bounds :duration 0.5)))
\end{minted}
\caption{An example of the presentations mechanism in Alloy to define the look and visual behaviour of a switch.}
\label{lst:presentation}
\end{listing}

@@

Particularly we had to write libraries to do the following:
\begin{enumerate}
\item Input handling for game controllers. This is rather involved, especially on Windows, where multiple APIs need to be supported simultaneously. Game controller devices are also prone to broken drivers or bad input mapping, which our library needs to deal with.
\item Audio output. We've implemented several different output APIs on both Linux and Windows, as both systems can differ on their supported output interfaces depending on version and setup.
\item Native dialog boxes. In order to show emergency error boxes or prompt the user for files.
\item Font discovery. To search the available fonts for a matching set we need to query operating system APIs.
\item Language querying. To ensure the game launches with the user's language (if supported of course), we again need to query the environment for the preferred localisation.
\end{enumerate}

@@

We make use of structures over standard objects only in very select cases where the impact of the increased verbosity can be contained, and the performance requirements actually indicated that we needed to optimise. Specifically this is for the hit information during collision resolution, and for the node graph that powers the pathing AI. In those cases the initial class approach produced too much overhead, so we switched to a few specific structures.

The dispatch overhead of CLOS has never been a real problem for us. Even with thousands of objects that receive events through our \code{handle} function, hundreds of methods attached to it, and tens of events every frame, not to mention all the other parts in the game that use CLOS, we still manage to easily hit a steady 120 frames per second on a ten year old machine.

@@

The first step in the deployment pipeline is the build of the actual game binary, for which we use ASDF and the Deploy library. With these all game code is compiled fresh, dumped into an executable suitable for the local platform, and bundled together with the depended-upon shared libraries, producing an executable ready for deployment to target machines.

>>>>

\author{Nicolas ``Shinmera'' Hafner}
\affiliation{%
  \institution{Shirakumo.org}
  \city{Zürich}
  \country{Switzerland}
}

@@

\begin{CCSXML}
  <ccs2012>
  <concept>
  <concept_id>10010147.10010371</concept_id>
  <concept_desc>Computing methodologies~Computer graphics</concept_desc>
  <concept_significance>300</concept_significance>
  </concept>
  <concept>
  <concept_id>10010520.10010570.10010574</concept_id>
  <concept_desc>Computer systems organization~Real-time system architecture</concept_desc>
  <concept_significance>300</concept_significance>
  </concept>
  <concept>
  <concept_id>10011007.10011006.10011066.10011070</concept_id>
  <concept_desc>Software and its engineering~Application specific development environments</concept_desc>
  <concept_significance>500</concept_significance>
  </concept>
  <concept>
  <concept_id>10011007.10011074.10011092.10011093</concept_id>
  <concept_desc>Software and its engineering~Object oriented development</concept_desc>
  <concept_significance>100</concept_significance>
  </concept>
  <concept>
  <concept_id>10011007.10011074.10011092.10011691</concept_id>
  <concept_desc>Software and its engineering~Error handling and recovery</concept_desc>
  <concept_significance>100</concept_significance>
  </concept>
  <concept>
  <concept_id>10011007.10011074.10011081.10011082</concept_id>
  <concept_desc>Software and its engineering~Software development methods</concept_desc>
  <concept_significance>100</concept_significance>
  </concept>
  </ccs2012>
\end{CCSXML}

\ccsdesc[300]{Computing methodologies~Computer graphics}
\ccsdesc[300]{Computer systems organization~Real-time system architecture}
\ccsdesc[500]{Software and its engineering~Application specific development environments}
\ccsdesc[100]{Software and its engineering~Object oriented development}
\ccsdesc[100]{Software and its engineering~Error handling and recovery}
\ccsdesc[100]{Software and its engineering~Software development methods}
\keywords{Common Lisp, Games, Video Games, Computer Graphics, Experience Report}

@@

All of the work we've done, including Trial, and even Kandria, is open source and available on our GitHub\cite{github}, in the hopes that it will inspire others to create new projects based upon them.

@@

Strandh\cite{strandh2014fast} proposes a new approach to improve the performance characteristics of generic function dispatch. Since we make heavy use of CLOS in our game, such improvements are very relevant.

Patton\cite{swcl-gc} are working on a new parallelisation of the SBCL garbage collector to allow for faster collection and thus reduced pause times. GC pause times are a frequently reported issue in games, as they can cause unexpected latency, and thus lead to drops in the framerate.

Mäkelä\cite{makela2021development} describe their experience in developing a game using the open source game engine Godot, with C\#. Godot is currently the leading open source game engine, and is striving to be a viable alternative to commercial products like Unity and Unreal.

Craighead et al.\cite{craighead2008using} cover a case study of using Unity, a proprietary game engine, to build a small game. Unity is currently most-used game engine for small to mid-sized games.

Nurminen\cite{nurminen1990rft} describe their experience developing and deploying a Common Lisp application to users.

@@

Usual desktop UI toolkits are rather cumbersome to style effectively, as most desktop applications are expected to present themselves in a ``native look and feel'', while games are expected to have rather elaborately customised and animated interfaces. Alloy's presentations protocol allows us to style the UI quite extensively. Thanks to macros we can present the user with a declarative style interface to define these behaviours with relative ease.

@@

Particularly we had to write libraries to do the following:
\begin{enumerate}
\item Input handling for game controllers. This is rather involved, especially on Windows, where multiple APIs need to be supported simultaneously.
\item Audio output. We've implemented several different output APIs on both Linux and Windows, as both systems can differ on their supported output interfaces depending on version and setup.
\item Native dialog boxes. In order to show emergency error boxes or prompt the user for files.
\item Font discovery. To search the available fonts for a matching set we need to query operating system APIs.
\item Language querying. To ensure the game launches with the user's language (if supported of course), we again need to query the environment for the preferred localisation.
\end{enumerate}

@@

We make use of structures over standard objects only in very select cases where the impact of the increased verbosity can be contained, and the performance requirements actually indicated that we needed to optimise. Specifically this is for the hit information during collision resolution, for the node graph that powers the pathing AI, and the glyph information used for text layouting and rendering. In those cases the initial class approach produced too much overhead with repeated slot access dispatch, so we switched to a few specific structures.

Especially with regards to propagating type information CLOS can be a hindrance, as you are prevented from declaring the argument and return types. Structures give us the advantage of fixing a return type on their accessors, propagating type information more easily and thus leading to better performing code at their call sites. There are projects that allow inlining method dispatch as well, eliminating some of the type ambiguity, but we have not seriously investigated these approaches yet.

Overall the dispatch overhead of CLOS has never been a real problem for us. Even with thousands of objects that receive events through our \code{handle} function, hundreds of methods attached to it, and tens of events every frame, not to mention all the other parts in the game that use CLOS, we still manage to easily hit a steady 120 frames per second on a ten year old machine.

@@

The first step in the deployment pipeline is the build of the actual game binary, for which we use the ASDF build system and the Deploy library. With these all game code is compiled fresh, dumped into an executable suitable for the local platform, and bundled together with the depended-upon shared libraries, producing an executable ready for deployment to target machines.
